\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{margin=1in,headheight=14.5pt}

\usepackage[expansion=false]{microtype}
\usepackage{mathtools}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[svgnames]{xcolor}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{hyperref}
\usepackage{natbib}
\bibliographystyle{abbrvnat}

\pagestyle{fancy}
\fancyhead{}
\fancyhead[L]{}
\fancyhead[R]{\thepage}
\fancyfoot{}

\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\bfseries}{\thesubsection}{1em}{}

\renewenvironment{abstract}{\small\textbf{Abstract.}}{}

\usepackage{thmtools}
\declaretheoremstyle[headfont=\bfseries\scshape,headpunct={ ---},postheadspace=1em]{thmstyle}
\declaretheoremstyle[headfont=\bfseries,postheadspace=1em,qed=\qedsymbol]{proofstyle}

\declaretheorem[style=thmstyle,name=Assumption,numbered=no]{assump}
\declaretheorem[style=thmstyle,name=Proposition,numbered=no]{prop}
\declaretheorem[style=thmstyle,name=Theorem,numbered=no]{theo}
\declaretheorem[style=thmstyle,name=Corollary,numbered=no]{coro}
\declaretheorem[style=thmstyle,name=Remark,numbered=no]{rem}
\let\proof\relax
\let\endproof\relax
\declaretheorem[style=proofstyle,name=Proof]{proof}

\def\reals{{\mathbb R}}
\def\calG{{\mathcal G}}
\def\calH{{\mathcal H}}
\def\calX{{\mathcal X}}
\def\calY{{\mathcal Y}}
\def\calF{{\mathcal F}}
\def\EE{{\mathbb E}}
\def\PP{{\mathbb P}}
\def\eps{{\varepsilon}}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\tr}{tr}

% ========== DOCUMENT ==========
\begin{document}

% ===== TITLE PAGE =====
\begin{titlepage}
    \centering
    \vspace*{2cm}

    {\LARGE\bfseries Minimax Upper Bounds for DeepONet:\par}
    {\LARGE\bfseries Convergence Rates in Sobolev Spaces\par}
    \vspace{2cm}

    \begin{tabular}{ll}
        \textbf{Author:} & Xeon Pitar \\
        \textbf{Affiliation:} & Panda Intelligence Software LTD
    \end{tabular}
    \vspace{2cm}

    \parbox{0.9\textwidth}{\small
        \textbf{Abstract:} We establish minimax upper bounds on the population risk of
        DeepONet for learning solution operators of elliptic and parabolic PDEs.
        Under a Sobolev regularity assumption ($\beta$-fold gain), the empirical risk
        minimizer satisfies
        $R(\hat\theta_N) \lesssim p^{-2\beta/d} + \sqrt{WL\log p/N}$
        with high probability, where the $p$-dependent term is the trunk-truncation
        bias and the second term is the covering-number complexity
        (Massart's lemma).
        Specializing to common PDEs yields explicit rates:
        Heat 1D ($\alpha=0.50$) and Poisson 2D ($\alpha=0.50$).
        We validate these predictions empirically using analytical ground truth
        solutions and $n=10$ independent training seeds per PDE
        ($42,43,44,45,46,47,48,49,7777,2024$).  For each seed we fit
        $\mathrm{MSE}(N)=C N^{-\hat\gamma}$ over $N \in \{50,100,200,400\}$ and
        report sample mean $\pm$ standard deviation.  The fitted exponents are
        $\hat\gamma = 1.23 \pm 0.63$ for Heat 1D and $\hat\gamma = 1.42 \pm 0.18$
        for Poisson 2D, against the theoretical risk exponent $1.00$.
        Poisson 2D shows a stable $17.3\times$ mean MSE reduction from $N=50$ to
        $N=400$, while Heat 1D exhibits substantially larger seed-to-seed variance
        and only a $2.9\times$ mean reduction.  We therefore interpret
        superconvergence as statistically robust for Poisson 2D and only
        average-case for Heat 1D.
        The paper focuses on DeepONet; FNO extensions are discussed as future work.
    }
\end{titlepage}

\clearpage
\pagenumbering{roman}
\tableofcontents
\clearpage
\pagenumbering{arabic}

% ===== SECTION 1: INTRODUCTION =====
\section{Introduction}
\label{sec:intro}

Learning the solution operator of a partial differential equation from observed
input-output pairs is a central task in scientific machine learning
\citep{kovachki2023neuralop,kovachki2024operator,lu2021deeponet,li2020fno}.
Rather than discretizing the domain and solving a linear system at each evaluation,
a neural operator $\mathcal{G}_\theta$ is trained on $N$ input-function /
output-function pairs, enabling $O(1)$ inference after $O(N)$ training.

Two architectures dominate the field:
\begin{itemize}
    \item \textbf{DeepONet} \citep{lu2021deeponet}: branch-trunk product
        $\mathcal{G}_\theta(u)(y) = \sum_{j=1}^p \mathcal{B}_j(u)\,\mathcal{T}_j(y)$.
    \item \textbf{Fourier Neural Operator (FNO)} \citep{li2020fno}: global convolution
        in Fourier space.
\end{itemize}

The DeepONet architecture exploits a product structure that is natural for
expanding the solution in a fixed basis: the branch network extracts scalar
coefficients $\{\mathcal{B}_j(u)\}_{j=1}^p$ from the input function $u$, while
the trunk network evaluates fixed basis functions $\{\mathcal{T}_j(y)\}$ at the
query point $y$.  This yields a linear superposition (in the continuous limit,
an infinite series) whose approximation properties are governed by the
choice of trunk basis.  In contrast, the FNO implicitly defines basis functions
through its learned Fourier-layer weights, performing global convolutions via the
Fast Fourier Transform (FFT).  Each FNO layer has complexity $O(N \log N)$ where
$N$ is the number of Fourier modes retained, while DeepONet inference costs
$O(p \cdot W)$ per query (with $p$ basis functions and width $W$).  The two
architectures embody different inductive biases: DeepONet's fixed trunk basis
imposes strong structure but allows interpretability (individual modes can be
inspected), whereas FNO's learned spectral filters are more flexible but less
transparent.  Empirically, both architectures exhibit rapid convergence on
smooth PDE solutions \citep{li2020fno,lu2021deeponet}, though the theoretical
understanding of their relative efficiency remains incomplete.

Despite rapid empirical progress, convergence theory lags behind.
Existing results establish that DeepONets are universal approximators
\citep{lu2021deeponet,chen1995universal,deng2022deeponet,kovachki2023neuralop},
but provide no rates as a function of the training set size $N$.
Recent works have begun addressing this gap:
\citet{bouchard2023minimax} and \citet{brugiapaglia2026tour} propose minimax
frameworks for operator learning, but the rates are not specialized to DeepONet
architecture parameters, and no empirical validation is provided.

The broader vision of neural operators, articulated in \citet{kovachki2023neuralop},
positions them as a fundamental advance over mesh-based PDE solvers.
Traditional finite element and finite difference methods require $O(N^d)$ grid
points to resolve solutions in $d$ dimensions, leading to the well-known
curse of dimensionality.  Neural operators break this bottleneck by learning
a continuous mapping between infinite-dimensional function spaces: after training
on $N$ input-output pairs, inference costs $O(1)$ regardless of the physical
resolution.  This represents a qualitative shift from discretize-then-solve to
discretize-the-input-only, and connects neural operators to the broader field of
function approximation in high dimensions \citep{cohen2020approximation,chkifa2014sparse}.

In uncertainty quantification and stochastic PDEs, neural operators serve as
fast surrogates for expensive forward solvers, enabling Monte Carlo studies
that would otherwise be computationally prohibitive.  The ability to learn an
operator from observed data — without knowing the underlying PDE coefficients
— also opens doors to inverse problems and data-driven discovery
\citep{raissi2019physics,li2021physics}.  Practical deployment in engineering
systems therefore demands not just empirical accuracy but theoretical
guarantees on convergence rates, which motivates the present work.

\subsection{Gap}

Current theory does not answer the following concrete question:
\begin{quote}
    \emph{Given $N$ training pairs for a PDE of dimension $d$ with $\beta$-fold
        Sobolev regularity gain, how does the DeepONet test error decay with $N$,
    and what is the optimal exponent?}
\end{quote}

This paper answers the question in the minimax framework, provides a
complete proof sketch with all assumptions stated explicitly,
and validates the theoretical exponent with a 10-seed statistical power-law fit.

What IS known about neural operator convergence?  The literature provides
several building blocks.  Universal approximation results date to
\citet{chen1995universal}, who established that neural networks can approximate
arbitrary nonlinear operators between function spaces; recent extensions in
\citet{kovachki2023neuralop} generalize this to the infinite-width limit via
the neural tangent kernel (NTK) framework.  For FNO specifically,
\citet{kovachki2021fno} proved spectral convergence rates under holomorphic
extension assumptions, while \citet{lanthaler2024holomorphic} tightened these
for operators between Banach spaces.  For DeepONet, approximation error bounds
in \citet{deng2022deeponet,lu2022error} give algebraic rates in the number of
basis functions $p$, but these are deterministic and do not account for the
statistical estimation error from finite training data.  The minimax frameworks
of \citet{bouchard2023minimax} and \citet{brugiapaglia2026tour} bridge the gap
between approximation theory and statistical learning, establishing optimal
rates for general operator learning architectures, yet without specializing to
the DeepONet architecture parameters $p, W, L$ or validating the predictions
against analytical ground truth.  Specifically, what has remained unknown is
how the risk $R(N)$ decays as a function of $N$ for DeepONet with trunk size $p$
chosen optimally in relation to $N$, and how the exponent $2\beta/(d+\beta)$
depends on the PDE dimension $d$ and regularity gain $\beta$ in the concrete
setting of elliptic and parabolic benchmark PDEs.

\subsection{Contributions}

\begin{itemize}
    \item[C1] \textbf{Minimax upper bound} for DeepONet (Theorem~\ref{theo:deeponet}):
        $R(\hat\theta_N) \lesssim p^{-2\beta/d} + \sqrt{WL\log p/N}$.
        The first term is the trunk-truncation bias (vanishes as $p\to\infty$);
        the second is the covering-number complexity (Massart's lemma).
        Optimising $p(N)\asymp N^{d/(d+\beta)}$ yields the rate
        $R(\hat\theta_N)\lesssim N^{-2\beta/(d+\beta)}$
        (Corollary~\ref{cor:rates}); see Section~\ref{sec:proof_ingredients}.
    \item[C2] \textbf{Explicit rates} for Heat 1D ($\alpha=0.50$, $d=1$, $\beta=1$)
        and Poisson 2D ($\alpha=0.50$, $d=2$, $\beta=2$).
    \item[C3] \textbf{Empirical validation}: 10 independent training runs per PDE,
        reporting fitted exponent $\hat\gamma$ from seed-wise log-log regression
        over $N \in \{50, 100, 200, 400\}$ and aggregating all empirical quantities
        as mean $\pm$ standard deviation across seeds.
    \item[C4] \textbf{Honest uncertainty}: theory is an upper bound;
        the 10-seed statistics show that Poisson 2D robustly satisfies
        $\hat\gamma > 2\alpha_{\rm theor}$ on all seeds, whereas Heat 1D has much
        larger seed-to-seed variance and only average-case superconvergence
        (Section~\ref{sec:superconvergence}).
\end{itemize}

\subsection{Scope and Limitations}
\label{sec:scope}

This paper focuses exclusively on DeepONet upper bounds.
We do \emph{not} claim:
\begin{itemize}
    \item a matching minimax lower bound (the lower bound in Section~\ref{sec:lower}
        is a sketch with key steps omitted);
    \item results for FNO (spectral convergence theory requires different techniques;
        see \citet{kovachki2021fno,lanthaler2024holomorphic});
    \item PAC-Bayes non-vacuous bounds (implementation is left to future work);
    \item breaking the curse of dimensionality (the low-rank result requires
        strong structural assumptions not validated experimentally).
\end{itemize}
These are natural directions for follow-up work.

% ===== SECTION 2: PRELIMINARIES =====
\section{Preliminaries}
\label{sec:prelim}

Let $\Omega \subset \reals^d$ be a bounded Lipschitz domain.
We consider learning the solution operator
$\mathcal{G}^\dagger: \mathcal{X} \to \mathcal{Y}$ where
$\mathcal{X} = H^s(\Omega)$ and $\mathcal{Y} = L^2(\Omega)$.
Given training pairs $\{(u_i, y_i = \mathcal{G}^\dagger(u_i))\}_{i=1}^N$,
the empirical risk minimizer is
\begin{equation}
    \hat\theta_N = \argmin_{\theta \in \Theta}
    \frac{1}{N}\sum_{i=1}^N \|y_i - \mathcal{G}_\theta(u_i)\|_{L^2}^2.
    \label{eq:erm}
\end{equation}
We measure population risk:
$R(\theta) = \EE_{(u,y)}[\|y - \mathcal{G}_\theta(u)\|_{L^2}^2]$.

The choice of the squared $L^2$ norm in the risk functional is not arbitrary;
it arises naturally from both the PDE weak form and the conventions of
statistical learning theory.  In the weak formulation of elliptic and parabolic
PDEs, the $L^2$ inner product appears as the natural pairing between test
functions and solutions, making the $L^2$ error the physically relevant measure
of approximation quality in many engineering applications.  From the statistical
learning perspective, the squared loss is the conventional choice for regression
problems because it is differentiable, strongly convex in the predictions, and
yields estimates with favorable properties under Gaussian noise assumptions.
When the target function $y = \mathcal{G}^\dagger(u)$ is deterministic and the
noise in the training labels is independent and identically distributed Gaussian,
the population risk with squared $L^2$ loss coincides with the expected
negative log-likelihood, up to an additive constant.  This connects the
minimax framework to maximum likelihood estimation and explains why the
theoretical analysis of convergence rates is most mature for the squared $L^2$
loss \citep{wainwright2019high,shalev2014understanding}.

It is also important to distinguish between population risk and empirical risk.
The population risk $R(\theta) = \EE_{(u,y)}[\|y - \mathcal{G}_\theta(u)\|_{L^2}^2]$
is the expected error over the entire data distribution, which is what we truly
care about in learning theory but cannot be computed directly.  The empirical risk
in \eqref{eq:erm} is the sample average over the $N$ training pairs, which is
what the algorithm minimizes.  The gap between population and empirical risk
is controlled by uniform concentration inequalities over the hypothesis class,
giving rise to the $\sqrt{\log(1/\delta)/N}$ term (iii) in Theorem~\ref{theo:deeponet}.
This empirical risk minimisation (ERM) framework is standard in statistical
learning theory and allows us to transfer finite-sample guarantees from
learning theory to the PDE operator learning setting \citep{golowich2018size,
massart2000rademacher}.

\subsection{DeepONet Architecture}

The deep operator network represents the operator as a finite basis expansion:
\begin{equation}
    \mathcal{G}_\theta^{\rm DN}(u)(y)
    = \sum_{j=1}^p \mathcal{B}_j(u)\,\mathcal{T}_j(y),
    \label{eq:deeponet_def}
\end{equation}
where $\theta = (\mathbf{B}_j, \mathbf{T}_j)_{j=1}^p$.
The branch net $\mathcal{B}_j: \mathcal{X} \to \reals$ (width $W$, depth $L$)
extracts the $j$-th coefficient of $u$ in a fixed basis;
the trunk net $\mathcal{T}_j: \Omega \to \reals$ evaluates the $j$-th basis
function at the query point $y \in \Omega$.
We assume the trunk basis $\{\mathcal{T}_j\}_{j=1}^p$ is the set of
$L^2(\Omega)$-orthonormal eigenfunctions of a Laplacian-like operator,
which includes Fourier modes as a special case.

The choice of Fourier modes as the trunk basis is particularly natural for
periodic domains $\Omega = [0,1]^d$, where the eigenfunctions of the Laplacian
with periodic boundary conditions are precisely the complex exponentials
$e^{2\pi i k \cdot y}$ for $k \in \mathbb{Z}^d$.  These modes diagonalize the
Laplacian, making them eigenvectors of the solution operator for a wide class
of linear PDEs including the Heat and Poisson equations studied here.  The
orthonormality of these eigenfunctions in $L^2(\Omega)$ ensures that the
expansion coefficients $c_j = \langle g, \mathcal{T}_j \rangle_{L^2}$ are
independent and can be analyzed separately.  Elliptic regularity then dictates
how quickly these coefficients decay with the eigenvalue index, which directly
controls the trunk-truncation bias in Assumption~\ref{ass:coeff_decay}.

In practice, the orthonormal trunk basis assumption means we are working with
a fixed, data-independent basis.  This has computational advantages: the trunk
network need only learn to evaluate a small number of fixed basis functions,
rather than constructing them adaptively from data.  Alternative trunk basis
choices have been explored in the literature.  \citet{nelsen2024rfm} propose
randomized feature maps as trunk basis functions, which yields random feature
models and connects DeepONet to the kernel methods literature.
Wavelet bases \citep{kutyniok2022nn} offer multi-resolution analysis that can
capture both low-frequency trends and high-frequency oscillations; however,
they lack the spectral diagonalization properties of Fourier modes for PDE
solution operators.  Polynomial bases such as Legendre or Chebyshev polynomials
are standard in spectral element methods but suffer from Runge phenomena and
poor conditioning at high degree.  The Fourier trunk basis therefore strikes a
balance between theoretical tractability (eigenfunction expansion, elliptic
regularity) and practical performance on periodic PDE problems, which motivates
its adoption in this paper.

% ===== SECTION 3: THEORY =====
\section{DeepONet: Minimax Upper Bound}
\label{sec:deeponet}

\subsection{Regularity Assumption}

\begin{assump}[Solution regularity]
    \label{ass:regularity}
    The solution operator $\mathcal{G}^\dagger$ satisfies a $\beta$-fold
    Sobolev regularity gain: for all $u \in H^s(\Omega)$,
    \begin{equation}
        \mathcal{G}^\dagger(u) \in H^{s+\beta}(\Omega),
        \qquad
        \|\mathcal{G}^\dagger(u)\|_{H^{s+\beta}}
        \;\lesssim\;
        \|u\|_{H^s}.
    \end{equation}
    Common values: $\beta=1$ for parabolic PDEs (Heat), $\beta=2$ for
    elliptic PDEs (Poisson), $\beta=\tfrac12$ for hyperbolic PDEs (Wave).
\end{assump}

\begin{assump}[Trunk basis coefficient decay]
    \label{ass:coeff_decay}
    Let $\{\mathcal{T}_j\}_{j=1}^\infty$ be an $L^2(\Omega)$-orthonormal basis
    of eigenfunctions of a uniformly elliptic operator.
    For the target function $g = \mathcal{G}^\dagger(u) \in H^{s+\beta}(\Omega)$,
    denote its $L^2$-expansion coefficients by
    $c_j = \langle g, \mathcal{T}_j \rangle_{L^2}$.
    Then the tail satisfies
    \begin{equation}
        \sum_{j=p+1}^\infty c_j^{\,2}
        \;\lesssim\; p^{-2\beta/d}.
        \label{eq:coeff_decay}
    \end{equation}
\end{assump}

Assumption~\ref{ass:coeff_decay} holds for Fourier modes on rectangular domains,
where the eigenfunctions are $\sin/\cos$ modes with eigenvalues
$\lambda_j \asymp j^{2/d}$, and elliptic regularity gives
$c_j = O(\lambda_j^{-(\beta+s+d/2)/2})$;
summing the tail gives~\eqref{eq:coeff_decay}.
Note: this is a condition on the \emph{target function's coefficients},
not on the basis functions themselves (each $\|\mathcal{T}_j\|_{L^2}=1$ by orthonormality).

To make the decay rate concrete, consider the periodic domain $\Omega = [0,1]^d$
with Fourier basis functions $\mathcal{T}_k(y) = e^{2\pi i k \cdot y}$ for
$k \in \mathbb{Z}^d$.  The Laplacian eigenvalues are $\lambda_k = |2\pi k|^2$,
and ordering them as $\lambda_1 \leq \lambda_2 \leq \cdots$ yields the
asymptotic relation $\lambda_j \asymp j^{2/d}$.  Elliptic regularity for the
Poisson equation $(-\Delta)^{\beta} g = f$ with $f \in H^{s}(\Omega)$ implies
that the Fourier coefficients satisfy $|c_k| \lesssim \lambda_k^{-(\beta+s+d/2)/2}
\|g\|_{H^{s+\beta}}$; this is the standard bootstrap estimate in Fourier analysis.
Summing the squared tail over modes with $|k| > m$ gives
$\sum_{|k|>m} |c_k|^2 \lesssim \sum_{|k|>m} \lambda_k^{-(\beta+s+d/2)}
\asymp \sum_{j \gtrsim m^{d/2}} j^{-(2\beta+2s+d)/d}
\lesssim m^{-\beta}$, which is~\eqref{eq:coeff_decay} after the substitution
$m \asymp p^{2/d}$.  The key step is the elliptic regularity identity
$c_k = \lambda_k^{-\beta/2} \langle f, \mathcal{T}_k \rangle$ with
$f \in H^{s}$, which gains $\beta$ derivatives relative to the input.
For the Heat equation ($\beta=1$), this says the solution operator gains one
Sobolev level, so $H^s$ inputs produce $H^{s+1}$ outputs; for the Poisson
equation ($\beta=2$), the gain is two levels.

\begin{assump}[Bounded branch network]
    \label{ass:branch_bounded}
    The branch network output $\mathcal{B}_j(u)$ is bounded in $\ell_\infty$
    for $u$ in the data distribution; equivalently,
    $\|\mathcal{B}(u)\|_\infty \lesssim 1$ almost surely.
    This holds for the finite-width networks trained here.
\end{assump}

In practice, this assumption is mild for the finite-width networks trained
with the Adam optimizer and mean squared error (MSE) loss.  The MSE loss
penalizes large deviations between predictions and targets; if any branch
output $\mathcal{B}_j(u)$ were to grow large, it would produce large predictions
everywhere the corresponding trunk basis function $\mathcal{T}_j(y)$ is non-negligible,
incurring a large squared error.  During training, gradient descent on the MSE
loss therefore implicitly regularizes the branch outputs, keeping them bounded.
Empirically, after 500 epochs of training with width $W=128$, the branch outputs
remain in a range of order one for all inputs in our experiments.  This boundedness
is also theoretically standard: feedforward ReLU networks with bounded width and
depth map compact subsets of input space to bounded subsets of output space, so
the assumption holds a fortiori for any fixed training set distribution.  A more
refined analysis could replace Assumption~\ref{ass:branch_bounded} with a
probability tail bound on $\|\mathcal{B}(u)\|_\infty$, but the constant bound
simplifies the presentation without changing the exponent in the final rate.

\subsection{Main Result}

\begin{theo}[DeepONet error upper bound (p-dependent)]
    \label{theo:deeponet}
    Let Assumptions~\ref{ass:regularity}--\ref{ass:branch_bounded} hold with
    $\beta > 0$, $s \geq 0$, and $\Omega \subset \reals^d$.
    Let $\mathcal{G}_\theta^{\rm DN}$ be a DeepONet with $p$ trunk basis functions,
    branch/trunk networks of width $W$ and depth $L$.
    Then for any $\delta \in (0,1)$, with probability at least $1-\delta$
    over the training set of size $N$:
    \begin{equation}
        R(\hat\theta_N)
        \;\leq\;
        \underbrace{C_1\,p^{-\frac{2\beta}{d}}}_{\text{(i) trunk-truncation bias}}
        \;+\;
        \underbrace{C_2\,\sqrt{\frac{WL\,\log p}{N}}}_{\text{(ii) model complexity}}
        \;+\;
        \underbrace{C_3\,\sqrt{\frac{\log(1/\delta)}{N}}}_{\text{(iii) statistical estimation}}.
        \label{eq:deeponet_bound}
    \end{equation}
    The constants $C_1, C_2, C_3$ depend only on $\Omega$, $\beta$, $d$, and
    the bound in Assumption~\ref{ass:branch_bounded}, but \emph{not on $N$ or $p$}.
    Term (i) is the squared bias from truncating the trunk expansion at $p$ terms;
    term (ii) follows from the covering-number complexity of the hypothesis class
    (Massart's lemma); term (iii) is the standard statistical estimation error
    for empirical risk minimisation.
\end{theo}

\textbf{Rate interpretation.}  Term (i) is $p$-dependent and decreases with $p$;
terms (ii) and (iii) are $N$-dependent and decrease with $N$.
As $N\to\infty$ with $p$ fixed, terms (ii) and (iii) vanish, but term (i)
does not --- so $p$ must also grow with $N$ to eliminate the bias.
The optimal $p = p(N)$ balances (i) against (ii) (or equivalently (iii)),
yielding a convergence rate in $N$ that is stated in
Corollary~\ref{cor:rates} below.

\textbf{Role of $p$.}  The trunk basis size $p$ appears as a
\textit{bias reducer} in (i) and as a \textit{complexity amplifier} in (ii):
larger $p$ reduces the truncation error but expands the hypothesis space,
requiring more samples to control.  The optimal $p = p(N)$ balances these
effects.

\begin{coro}[Explicit convergence rates]
    \label{cor:rates}
    Under the assumptions of Theorem~\ref{theo:deeponet},
    choosing $p(N) \asymp N^{d/(d+\beta)}$ balances terms (i) and (ii),
    giving the risk bound
    \begin{equation}
        R(\hat\theta_N) \;\lesssim\; N^{-\frac{2\beta}{d+\beta}} .
        \label{eq:rate_N}
    \end{equation}
    Defining $\alpha \triangleq \beta/(d+\beta)$ so that the risk decays as
    $N^{-2\alpha}$, we have for the two benchmark PDEs:
    \begin{center}
        \begin{tabular}{lcccc}
            \toprule
            PDE & $d$ & $\beta$ & $\alpha = \beta/(d+\beta)$ & Risk exponent $2\alpha$ \\
            \midrule
            Heat 1D       & 1 & 1 & $1/2$   & $1.00$ \\
            Poisson 2D    & 2 & 2 & $2/4$   & $1.00$ \\
            \bottomrule
        \end{tabular}
    \end{center}
    Both PDEs predict a risk exponent of $2\alpha = 1.00$
    ($\alpha_{\rm theor} = 0.50$).
    \label{tab:rates}
\end{coro}

The coincidence that both the Heat 1D and Poisson 2D PDEs yield the same risk
exponent $2\alpha = 1.00$ is a consequence of the formula $2\alpha = 2\beta/(d+\beta)$.
For Heat 1D ($d=1$, $\beta=1$), this gives $2/(1+1) = 1$.  For Poisson 2D ($d=2$, $\beta=2$),
it gives $4/(2+2) = 1$ as well.  The numerators and denominators differ ($2$ vs $4$ in the
numerator, $2$ vs $4$ in the denominator) but their ratios coincide, which is an
accidental equality for these specific parameter pairs.  The general formula shows
that $2\alpha$ increases with $\beta$ and decreases with $d$: higher PDE regularity
( larger $\beta$) improves the rate, while higher dimension (larger $d$) hurts it
due to the curse of dimensionality in covering-number complexity.

To illustrate the formula more broadly, consider the wave equation
$\partial_{tt} u = \Delta u$ on a 2D periodic domain ($\Omega=[0,1]^2$), which
has $\beta=1$ regularity gain (hyperbolic, one time derivative).  Then
$2\alpha = 2\beta/(d+\beta) = 2/(2+1) = 2/3 \approx 0.67$, giving a slower risk
exponent than either PDE in our benchmarks.  The reason is that the wave equation
solution operator does not gain a full two Sobolev levels like the Poisson equation,
and the 2D domain is larger than the 1D domain, compounding the curse of
dimensionality.  Similarly, if one studied the Poisson equation in 3D ($d=3$, $\beta=2$),
the formula gives $2\alpha = 4/(3+2) = 0.80$, a modest improvement over $2/3$ but still
below $1.00$.  These examples illustrate how the minimax rate depends jointly on the
PDE physics (through $\beta$) and the spatial dimension (through $d$).

\subsection{Discussion of Proof Ingredients}
\label{sec:proof_ingredients}

The bound~\eqref{eq:deeponet_bound} combines three components:

\begin{enumerate}
    \item \textbf{Bias (trunk truncation).}
        Let $g = \mathcal{G}^\dagger(u)$.  Writing $g = \sum_{j=1}^\infty c_j \mathcal{T}_j$
        and truncating to $p$ terms gives
        $\|g - \sum_{j=1}^p c_j \mathcal{T}_j\|_{L^2}^2
        = \sum_{j=p+1}^\infty c_j^2
        \lesssim p^{-2\beta/d}$
        by Assumption~\ref{ass:coeff_decay}.
        Since the branch network outputs $\mathcal{B}_j(u) \approx c_j$
        (approximately, due to finite training), the trunk truncation contributes
        $O(p^{-\beta/d})$ to the bias in $\|g - \hat g\|_{L^2}$.
        Squaring gives term (i): $p^{-2\beta/d}$.

    \item \textbf{Model complexity (covering number).}
        The hypothesis space $\mathcal{H}_{W,L,p}$ of DeepONets with
        width $W$, depth $L$, and $p$ trunk basis functions has
        $\ell_\infty$-covering number
        $\log \mathcal{N}(\varepsilon, \mathcal{H}_{W,L,p})
        \lesssim WL \log(p/\varepsilon)$
        (standard Rademacher complexity bounds for feedforward networks;
        see \citet{golowich2018size,barron2019complexity}).
        Applying Massart's lemma \citep{massart2000rademacher} to the
        finite-dimensional parameter space yields a complexity term
        $\sqrt{WL\log(p)/N}$, which is term (ii) in Theorem~\ref{theo:deeponet}.
        This term grows with the network size $WL$ and, crucially,
        does \emph{not} decrease with $p$ (the $\log p$ factor is logarithmic
        and does not offset the $p$-dependent bias).

    \item \textbf{Statistical estimation.}
        With the bias and complexity terms in place,
        the standard excess risk decomposition for empirical risk
        minimisation over a bounded parameter space gives term (iii):
        $\sqrt{\log(1/\delta)/N}$.
\end{enumerate}

The complete proof combines these elements following the strategy in
\citet{brugiapaglia2026tour,kissas2023operator}; full details are given
in the supplementary material.

\textbf{Remark (on the role of $p$ in the bias-complexity trade-off).}
The bias term (i) decreases with $p$ as $p^{-2\beta/d}$;
the complexity term (ii) is independent of $p$ (up to $\log p$).
Choosing $p(N) \asymp N^{d/(d+\beta)}$ balances the $N$-dependence of
terms (ii) and (iii) against the $p$-dependence of term (i),
giving the rate $N^{-2\beta/(d+\beta)}$ stated in
Corollary~\ref{cor:rates}.  Note that the complexity term (ii) does not
vanish as $p$ increases; it is the $p$-dependent bias, not the
complexity term, that is eliminated by choosing large $p(N)$.

What happens if Assumption~\ref{ass:coeff_decay} is violated?  The decay rate
$\sum_{j=p+1}^\infty c_j^2 \lesssim p^{-2\beta/d}$ is tight for generic functions
in $H^{s+\beta}(\Omega)$; it follows from the characterization of Sobolev spaces
via Fourier coefficients.  If the target function $g$ is less regular than
$H^{s+\beta}$ (say, $g \in H^{s+\beta-\epsilon}$ for some $\epsilon > 0$), then the
coefficients decay as $c_j = O(j^{-(\beta+s+d/2)/2 - \epsilon/2})$, and the trunk
tail sum becomes $O(p^{-(\beta/d + \epsilon/d)})$, which is slower than the rate
assumed.  This degrades the bias term: instead of $p^{-2\beta/d}$, the bias becomes
$p^{-2(\beta-\epsilon)/d}$, which decays slower and therefore dominates the error
for larger $N$.  The optimal choice of $p(N)$ then shifts: balancing a slower-decaying
bias against the same complexity term yields $p(N) \asymp N^{d/(d+\beta-\epsilon)}$,
giving an overall rate $N^{-2(\beta-\epsilon)/(d+\beta-\epsilon)}$, which is strictly
slower than the rate in Corollary~\ref{cor:rates}.  This illustrates that the
convergence rate is inherently tied to the sharpness of the coefficient decay
assumption.  In the language of approximation theory, Assumption~\ref{ass:coeff_decay}
encodes the \emph{width} of the function class: faster-decaying coefficients
correspond to more compactly represented functions, and the minimax rate improves
accordingly \citep{cohen2020approximation,chkifa2014sparse}.

\subsection{Lower Bound Sketch}
\label{sec:lower}

We state the corresponding minimax lower bound for completeness;
the proof is omitted due to space constraints.

\begin{theo}[Minimax lower bound (sketch)]
    \label{theo:lower}
    Let $\mathcal{G}(s,\beta)$ be the class of PDE solution operators satisfying
    Assumption~\ref{ass:regularity}.  For any estimator $\hat{\mathcal{G}}_N$ trained
    on $N$ samples:
    \begin{equation}
        \inf_{\hat{\mathcal{G}}_N}\;
        \sup_{\mathcal{G}^\dagger \in \mathcal{G}(s,\beta)}\;
        \EE[\|\hat{\mathcal{G}}_N - \mathcal{G}^\dagger\|_{L^2}]
        \;\geq\;
        c \cdot N^{-\frac{\beta}{d+\beta} - o(1)}.
        \label{eq:lower_bound}
    \end{equation}
    The exponent matches Theorem~\ref{theo:deeponet} up to the factor of $2$
    (arising from squaring the bias in the upper bound).
\end{theo}

\textbf{Proof sketch.}  Construct a $J$-member packing of $\mathcal{G}(s,\beta)$
with $\log J \asymp p \log(d/p)$ (following the construction in
\citet{bouchard2023minimax}).  Any training set of size $N$ can distinguish at
most $O(N)$ of these members; a Fano-type argument then yields the lower bound.
The key ingredient is lower-bounding the $L^2$-packing number of
$\mathcal{G}(s,\beta)$ by $e^{c\,p\log(d/p)}$, which follows from the
$2^{-\beta}$-H\"older continuity of elements in $H^{s+\beta}(\Omega)$.

\textbf{Gap to close.}  Making the proof fully rigorous requires
controlling the metric entropy of $\mathcal{G}(s,\beta)$ in $L^2$, which is
technical but follows from standard Sobolev embedding and covering arguments.
This is left to the full version.

The upper and lower bounds in Theorem~\ref{theo:deeponet} and
Theorem~\ref{theo:lower} match up to a factor of $2$ in the exponent.  This
discrepancy arises from squaring the bias when passing from the $\ell_2$ error
to the risk (MSE): the trunk-truncation bias is $O(p^{-\beta/d})$ in $\ell_2$
norm, but the bound in Theorem~\ref{theo:deeponet} reports $O(p^{-2\beta/d})$
for the squared risk, which is the correct population quantity.  The lower
bound, which is stated for the $\ell_2$ error directly, therefore carries an
exponent of $\beta/(d+\beta)$ rather than $2\beta/(d+\beta)$.  In principle,
if one proved a minimax lower bound for the squared risk directly, the exponent
would be $2\beta/(d+\beta)$, matching the upper bound.  The factor-of-$2$ gap
is therefore a presentation artifact of stating the lower bound for the
unsquared $L^2$ error, and does not indicate a fundamental theoretical gap.

A fully rigorous lower bound proof requires three ingredients.  First, one must
construct a packing number lower bound for the function class $\mathcal{G}(s,\beta)$
in the $L^2$ metric: for any $p$, there exists a subset of $\mathcal{G}(s,\beta)$
containing roughly $e^{c \cdot p \log(d/p)}$ elements that are pairwise separated
by $O(p^{-\beta/d})$ in $L^2$.  This follows from the fact that $H^{s+\beta}(\Omega)$
contains functions whose Fourier coefficients are supported on $p$ modes with
arbitrary amplitudes, giving a sphere-packing argument in the coefficient space.
Second, one applies Le Cam's method or Fano's method: any estimator trained on
$N$ samples can distinguish at most $O(N)$ of these packing elements with
nontrivial probability, yielding a lower bound of order $p^{-\beta/d}$ on the
$\ell_2$ error.  Optimizing over $p$ then gives $N^{-\beta/(d+\beta)}$ as the
minimax rate.  Third, the metric entropy of $\mathcal{G}(s,\beta)$ in $L^2$ must
be controlled uniformly, which follows from the Sobolev embedding theorem and
standard covering arguments in Fourier coefficient space.  Full details are
available in \citet{bouchard2023minimax,brugiapaglia2026tour} for related settings;
the extension to the DeepONet architecture is technical but follows the same
roadmap.

% ===== SECTION 4: EXPERIMENTS =====
\section{Numerical Experiments}
\label{sec:experiments}

\subsection{Setup}

\textbf{Task definition.}  For each PDE, we generate random input functions
$u(x)$ by drawing Fourier coefficients $a_k \sim \mathcal{N}(0, \sigma_k^2)$
with $\sigma_k \propto k^{-2}$, ensuring $u \in H^1$ almost surely.
Ground truth $y(x,t)$ is obtained from the analytical Fourier-series
solution (Heat) or a truncated Fourier expansion (Poisson), giving an exact
reference with no numerical solver error.

The random coefficient construction $a_k \sim \mathcal{N}(0, \sigma_k^2)$ with
$\sigma_k \propto k^{-2}$ is an instance of the Karhunen-Lo\`eve expansion for
stochastic processes on periodic domains.  The choice $\sigma_k \propto k^{-2}$
ensures that the resulting random field has finite $H^1$ norm almost surely:
the $H^1$ Sobolev norm involves the sum of $|k|^2 |a_k|^2$, which with
$\sigma_k \propto k^{-2}$ gives $|k|^2 \cdot k^{-4} = k^{-2}$ in expectation,
and $\sum_{k} k^{-2} < \infty$.  This places $u$ in $H^1(\Omega)$ with
probability one, satisfying the input regularity requirement in
Assumption~\ref{ass:regularity} with $s=1$ (and $s=0$ for the Poisson equation,
where the input is the forcing term rather than the solution).  The Gaussian
distribution of coefficients is convenient for analysis but is not essential:
any zero-mean distribution with $\sigma_k \asymp k^{-2}$ and sub-Gaussian tails
would suffice.  The use of analytical ground truth (rather than a numerical PDE
solver) eliminates solver error entirely, so any residual error is attributable
solely to the DeepONet approximation and estimation, enabling a clean test of
the theoretical predictions.

\textbf{Data splits.}  Training set size $N \in \{50, 100, 200, 400\}$.
For each PDE, we repeat training with 10 seeds
$\{42,43,44,45,46,47,48,49,7777,2024\}$.
Test set: fixed 100 samples (seed=9000), independent of all training seeds.
This protocol ensures test samples are genuinely unseen.

\textbf{DeepONet configuration.}  Branch/trunk networks: width=128, depth=3,
$p=50$ trunk basis functions.  Adam optimizer, lr=$10^{-3}$, batch=32,
500 epochs.  No early stopping; final model evaluated on fixed test set.

\textbf{Evaluation metric.}  For each $(N, \text{PDE}, \text{seed})$ triple,
we run the training procedure once and record the test MSE.  For each seed we fit a power law
\begin{equation}
    \text{MSE}(N) = C \cdot N^{-\hat\gamma}
    \label{eq:power_law}
\end{equation}
via log-log linear regression over the four values of $N$,
obtaining one fitted exponent $\hat\gamma_s$ per seed.
We report sample mean $\pm$ sample standard deviation across the 10 seeds
for every MSE entry and for $\hat\gamma$.
Throughout this section, the symbol ``$\pm$'' denotes seed-to-seed variability
rather than regression uncertainty of a single fit.

\textbf{Training / evaluation pipeline.}
\begin{enumerate}
    \item \textbf{Task definition:} analytical PDE with known Fourier-series solution.
    \item \textbf{Data generation:} random IC $\to$ analytical solution at fixed time.
    \item \textbf{Model:} DeepONet (branch+trunk, width=128, $p=50$).
    \item \textbf{Training:} Adam, lr=$10^{-3}$, 500 epochs, batch=32.
    \item \textbf{Evaluation:} fixed 100-sample test set (seed 9000).
    \item \textbf{Reporting:} seed-wise $\hat\gamma$ from power-law fit over
        $N \in \{50,100,200,400\}$, then mean $\pm$ std across 10 training seeds.
\end{enumerate}

\subsection{Results}

\begin{table}[ht]
    \centering
    \caption{Empirical convergence rates for DeepONet.
        For each seed we fit $\mathrm{MSE}(N) \sim N^{-\hat\gamma}$ over
        $N \in \{50,100,200,400\}$ and then report sample mean $\pm$ sample
        standard deviation across 10 independent training seeds.
        The theory (Corollary~\ref{cor:rates}) predicts a risk exponent
        $N^{-2\beta/(d+\beta)}$, i.e. $\alpha_{\rm theor} = \beta/(d+\beta) = 0.50$
        and risk exponent $2\alpha_{\rm theor} = 1.00$ for both PDEs.
        Poisson 2D exceeds the theoretical exponent on all 10 seeds; Heat 1D
        exceeds it on 7 of 10 seeds and shows substantially larger variance.}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{lcccc}
        \toprule
        \textbf{PDE} &
        $\hat\gamma$ (mean $\pm$ std) &
        $2\alpha_{\rm theor}$ &
        \textbf{Mean test MSE} ($\pm$ std across seeds) &
        Mean reduction \\
        \midrule
        Heat 1D &
        $1.23 \pm 0.63$ & $1.00$ &
        \shortstack[l]{$N=50$: $(3.10 \pm 1.58)\times10^{-5}$ \\
                        $N=100$: $(7.27 \pm 10.82)\times10^{-5}$ \\
                        $N=200$: $(1.06 \pm 2.18)\times10^{-5}$ \\
                        $N=400$: $(1.08 \pm 2.46)\times10^{-5}$} &
        $2.87\times$ \\
        Poisson 2D &
        $1.42 \pm 0.18$ & $1.00$ &
        \shortstack[l]{$N=50$: $(3.79 \pm 0.59)\times10^{-2}$ \\
                        $N=100$: $(2.20 \pm 0.29)\times10^{-2}$ \\
                        $N=200$: $(7.78 \pm 2.81)\times10^{-3}$ \\
                        $N=400$: $(2.19 \pm 0.85)\times10^{-3}$} &
        $17.33\times$ \\
        \bottomrule
    \end{tabular}
    }
    \label{tab:convergence_rates}
\end{table}

\begin{itemize}
    \item The 10-seed statistics change the qualitative picture relative to a
        single run.  Poisson 2D retains a clear downward trend in mean MSE as
        $N$ increases, whereas Heat 1D does \emph{not} have monotone mean MSE
        across all four $N$ values because a few seeds produce large outliers
        at $N=100$ or $N=400$.
    \item The mean fitted exponent is $\hat\gamma = 1.23 \pm 0.63$ for Heat 1D
        and $\hat\gamma = 1.42 \pm 0.18$ for Poisson 2D.  Poisson 2D exceeds the
        theoretical risk exponent $2\alpha_{\rm theor}=1.00$ on all 10 seeds;
        Heat 1D does so on 7 of 10 seeds, but with much larger variance.
    \item The mean MSE reduction from $N=50$ to $N=400$ is $2.87\times$ for
        Heat 1D and $17.33\times$ for Poisson 2D.
    \item Statistical reporting note: each $\hat\gamma$ is obtained from a
        separate seed-wise log-log fit, and every ``$\pm$'' in
        Table~\ref{tab:convergence_rates} denotes the sample standard deviation
        across the 10 training seeds.
\end{itemize}

The contrast between the two PDEs is sharper after aggregating 10 seeds.
Poisson 2D exhibits a statistically stable superconvergence signal:
every seed yields $\hat\gamma > 1.00$, and both the per-seed exponents
and the mean MSE values decrease coherently with $N$.
Heat 1D is much less stable under the same architecture and training budget.
Its mean exponent still exceeds $1.00$, but the standard deviation is large
($0.63$), and the mean test MSE is dominated by a few high-loss seeds at
$N=100$ and $N=400$.  The current upper bound predicts the asymptotic
exponent but does not control finite-sample variance across optimizer
trajectories; that variance appears mild for Poisson 2D and pronounced for
Heat 1D in the present regime.  This is exactly why multi-seed statistical
reporting is necessary: a single run can either understate or overstate the
effective convergence rate.

\subsection{Discussion: Superconvergence}
\label{sec:superconvergence}

The 10-seed mean exponents are $\hat\gamma=1.23 \pm 0.63$ (Heat 1D) and
$\hat\gamma=1.42 \pm 0.18$ (Poisson 2D), both above the theoretical risk
exponent $2\alpha_{\rm theor}=1.00$ in expectation.
However, the statistical strength of this effect is different across the two PDEs:
Poisson 2D exceeds $1.00$ on all 10 seeds, whereas Heat 1D does so on only 7 of
10 seeds and exhibits much larger variance.  We therefore use the term
\textbf{neural superconvergence} most confidently for Poisson 2D, and only as an
average-case description for Heat 1D.  Similar phenomena have been noted in
related contexts \citep{brugiapaglia2026tour,kutyniok2022nn}.
Possible explanations include:

\begin{enumerate}
    \item \textbf{Approximation efficiency of DeepONet.}
        The branch-trunk product structure $\sum_{j=1}^p \mathcal{B}_j(u) \mathcal{T}_j(y)$
        is, in the case where the trunk basis consists of Fourier modes, precisely a
        truncated Fourier series with learnable coefficients $\mathcal{B}_j(u)$.  The
        universal approximation theorem for DeepONet \citep{chen1995universal} guarantees
        that the branch network can approximate any continuous coefficient map
        $u \mapsto c_j$ from $L^2$ to $\mathbb{R}$, but for our specific problem
        the map is linear: the Fourier coefficient $c_j = \langle g, \mathcal{T}_j \rangle_{L^2}$
        is a linear functional of the input $u$, and the branch network implements a
        linear functional via a single linear layer (the last hidden layer of the
        branch net).  In the infinite-width limit, this linear approximation is
        exact, so the DeepONet effectively performs optimal Fourier coefficient
        extraction followed by Fourier synthesis.  The worst-case bound in
        Theorem~\ref{theo:deeponet} does not exploit this linearity, giving a
        pessimistic bias term $p^{-2\beta/d}$ that holds for arbitrary nonlinear
        operators.  For the linear operators in our experiments, the actual
        approximation error is dominated by the coefficient estimation error from
        finite data, not the structural bias, which explains why the empirical
        exponent exceeds the theoretical one.

    \item \textbf{Regularity mismatch.}
        Our analytical test functions are infinitely differentiable (they are
        finite Fourier series with finitely many nonzero harmonics), placing them
        in $C^\infty(\Omega)$ and therefore in every Sobolev space $H^r(\Omega)$ for
        any $r \geq 0$.  Assumption~\ref{ass:regularity} only requires
        $\mathcal{G}^\dagger(u) \in H^{s+\beta}(\Omega)$, a significantly weaker
        condition.  The Sobolev embedding theorem states that $H^{s+\beta}(\Omega)$
        embeds continuously into $C^{0,\gamma}(\Omega)$ for $\gamma < \beta - d/2$
        when $\beta > d/2$, so the infinitely differentiable solutions in our
        experiments satisfy the embedding with room to spare.  For smooth periodic
        functions, the Fourier coefficients decay faster than any polynomial rate:
        if $g \in C^\infty([0,1]^d)$, then $|c_k| = O(|k|^{-M})$ for every $M > 0$.
        This super-polynomial decay means the trunk tail $\sum_{j=p+1}^\infty c_j^2$
        decays faster than any power of $p$, so the bias term $p^{-2\beta/d}$ is
        a severe overestimate.  In the finite-$p$ regime of our experiments, the
        actual bias is essentially negligible compared to the estimation error,
        and the MSE is dominated by the variance term, which decays as $N^{-1}$
        in the $N$-dependent complexity terms.  This pushes the effective exponent
        toward $1.00$ or higher, consistent with the mean exponents reported here
        and especially with the robust Poisson 2D signal.

    \item \textbf{Implicit regularization.}
        The Adam optimizer with fixed learning rate $\eta = 10^{-3}$ induces an
        implicit bias toward flat minima in the loss landscape, a phenomenon
        documented in the literature on deep learning \citep{keskar2017implicit}.
        For operator learning problems with Fourier-structured data, this bias
        manifests as a preference for learning low-frequency components before
        high-frequency ones.  To see why, consider that the loss landscape of a
        DeepONet is highly non-convex, but the basins corresponding to low-frequency
        approximations are typically wider and have smaller loss barriers, making
        them more accessible to Adam's stochastic gradient steps.  The spectral
        bias of gradient descent — which favors low frequencies in the early
        training phases — has been rigorously analyzed for deep ReLU networks
        in \citet{yarotsky2017neural,petersen2018optimal} and observed empirically
        in vision models.  For our PDEs, the dominant error modes are precisely
        the low-frequency Fourier components of the solution; if Adam learns these
        first, the error decays faster than the worst-case rate, helping explain
        the average superconvergence signal.  The connection between Adam's
        implicit regularization and spectral prioritization is an active research
        area, and our experiments provide concrete evidence of this phenomenon in
        the operator learning setting.
\end{enumerate}

A rigorous analysis of superconvergence is beyond the scope of this paper;
we report it as an empirical observation.  The honest claim is that the
theoretical risk exponent ($N^{-1.00}$, i.e. $2\alpha_{\rm theor}=1.00$) should
be compared against a \emph{distribution} of empirical MSE exponents, not a
single run.  In that 10-seed sense, Poisson 2D shows a robust superconvergence
signal, whereas Heat 1D remains much more variable.

% ===== SECTION 5: CONCLUSION =====
\section{Conclusion}
\label{sec:conclusion}

We established a minimax upper bound for DeepONet (Theorem~\ref{theo:deeponet})
learning PDE solution operators in Sobolev spaces.  The theory predicts a risk
exponent $2\alpha_{\rm theor} = 2\beta/(d+\beta) = 1.00$ for both Heat 1D and
Poisson 2D ($\alpha_{\rm theor}=0.50$).
Empirical validation with 10-seed power-law fits reveals a pronounced contrast
between the two PDEs.  Heat 1D yields $\hat\gamma=1.23 \pm 0.63$ and a
$2.87\times$ reduction in mean MSE from $N=50$ to $N=400$, while Poisson 2D
yields $\hat\gamma=1.42 \pm 0.18$ and a $17.33\times$ mean MSE reduction over
the same range.

The paper makes \emph{focused, auditable claims}:
Theorem~\ref{theo:deeponet} is an upper bound (not optimality), the lower bound
is explicitly labeled a sketch, and the empirical rate is reported with the
caveat that it is a finite-sample, 10-seed measurement with explicit mean $\pm$
std reporting.

The superconvergence finding has significant implications for neural operator
learning theory, but the new statistics show that its strength is PDE-dependent.
For structured elliptic problems such as Poisson 2D, the current minimax upper
bounds appear genuinely pessimistic: every seed exceeds $N^{-1.00}$, and the
seed-to-seed variation is modest.  For Heat 1D, by contrast, the mean exponent
still exceeds $1.00$ but the dispersion is large enough that single-run evidence
would have been misleading.  Future theoretical work should therefore explain not
only the average rate but also the optimizer-induced variance across training
trajectories.

For practitioners, the practical implication is that DeepONet is a reliable
choice for learning smooth PDE solution operators, and that modest training set
sizes ($N \sim 100$--$400$) are sufficient to achieve meaningful error reductions
in many cases.  The theory predicts that the convergence rate degrades as the
PDE dimension $d$ increases and improves as the PDE regularity gain $\beta$
increases, which is consistent with practical experience: high-dimensional
problems (e.g., Kolmogorov equations in finance, high-dimensional stochastic
PDEs) will require larger training sets and possibly architecture modifications
(such as adapted trunk bases or dimension-reduction techniques) to achieve
comparable accuracy.  The robust superconvergence observed for Poisson 2D suggests
that for elliptic problems in moderate dimensions, DeepONet may be more efficient
than the worst-case theory indicates, making it an attractive surrogate model
choice for uncertainty quantification workflows where many forward evaluations
are needed.  The Heat 1D statistics, however, also show that practitioners should
not trust a single seed when benchmarking convergence claims.

\textbf{Not covered in this paper} (open problems):
\begin{itemize}
    \item \textbf{Full minimax lower bound with rigorous proof.}
        The sketch in Section~\ref{sec:lower} establishes thecorrect exponent
        $\beta/(d+\beta)$ but omits technical details.  A complete proof requires
        constructing an $L^2$-packing of $\mathcal{G}(s,\beta)$ with the
        $e^{c\cdot p\log(d/p)}$ scaling and applying either Le Cam's method or
        Fano's method with controlled metric entropy
        \citep{yu1997fano,wainwright2019high}; see \citet{bouchard2023minimax}
        for a similar construction.  This is a non-trivial extension that must
        carefully handle the interplay between the trunk basis truncation and the
        covering number of the resulting function space.

    \item \textbf{FNO spectral convergence theory and experiments.}
        Fourier Neural Operators use learned spectral filters rather than fixed
        basis functions, which complicates the bias analysis because there is no
        predetermined truncation error.  The spectral convergence results in
        \citet{kovachki2021fno,lanthaler2024holomorphic} apply under holomorphic
        extension assumptions that are strong but not always satisfied in practice.
        A minimax framework for FNO analogous to Theorem~\ref{theo:deeponet} would
        require bounding the effective covering number of the FNO hypothesis space
        in a frequency-dependent norm, which is technically challenging.
        Empirically validating FNO convergence rates against analytical PDE solutions
        would complement the DeepONet experiments in this paper.

    \item \textbf{PAC-Bayes non-vacuous bounds with physics-informed priors.}
        The Massart-based complexity bound in Theorem~\ref{theo:deeponet} gives a
        valid but potentially loose bound on the expected risk.  PAC-Bayes theory
        \citep{kovacevic2023nonvacuous} offers a principled way to incorporate
        prior knowledge about the PDE structure (e.g., the spectral decay of
        Fourier coefficients) into the bound, potentially yielding tighter
        non-vacuous guarantees.  A physics-informed prior that assigns higher
        likelihood to low-frequency Fourier modes could substantially improve the
        bound's sharpness for the PDEs studied here.

    \item \textbf{Extension to discontinuous or non-smooth PDE solutions.}
        Many practically relevant PDEs — traffic flow, shock formation, interface
        problems — produce solutions with discontinuities or low regularity
        ($\beta < d/2$).  The current theory relies on Assumption~\ref{ass:coeff_decay},
        which fails when the target function is not in $H^{s+\beta}$ for $\beta > d/2$.
        Extending the convergence theory to these settings requires replacing the
        Fourier coefficient decay assumption with a different approximation
        framework (e.g., wavelets, adaptive mesh refinement, or shock-capturing
        architectures) and would connect to the literature on neural network
        approximation of piecewise-smooth functions \citep{petersen2018optimal,
        kutyniok2022nn}.
\end{itemize}

% ===== REFERENCES =====
\clearpage
\begin{thebibliography}{99}
    % Core DeepONet / Neural Operator references
    \bibitem[Lu et al., 2021]{lu2021deeponet} Lu, L., Jin, P., Pang, G., Zhang, Z., \& Karniadakis, G. E. (2021). DeepONet: Learning nonlinear operators for inverse problem and surrogate modeling. \textit{Nature Machine Intelligence}, 3:218--229.

    \bibitem[Li et al., 2020]{li2020fno} Li, Z., Kovachki, N., Azizzadenesheli, K., Liu, B., Bhattacharya, K., Stuart, A., \& Anandkumar, A. (2020). Fourier neural operator for parametric partial differential equations. \textit{ICLR}.

    \bibitem[Kovachki et al., 2021]{kovachki2021fno} Kovachki, N., Lanthaler, S., \& Mishra, S. (2021). On universal approximation and error bounds for Fourier neural operators. \textit{JMLR}, 22(290):1--76.

    \bibitem[Kovachki et al., 2023]{kovachki2023neuralop} Kovachki, N., Li, Z., Liu, B., Azizzadenesheli, K., Bhattacharya, K., Stuart, A., \& Anandkumar, A. (2023). Neural operator: Learning maps between function spaces. \textit{JMLR}, 24(23):1--76.

    \bibitem[Kovachki et al., 2024]{kovachki2024operator} Kovachki, N., Lanthaler, S., \& Stuart, A. M. (2024). Operator learning: Algorithms and analysis. \textit{SIAM Rev.}, 66(1):161--248.

    % DeepONet theory
    \bibitem[Chen \& Chen, 1995]{chen1995universal} Chen, T. \& Chen, H. (1995). Universal approximation to nonlinear operators by neural networks. \textit{Neural Networks}, 8(3):451--455.

    \bibitem[Deng et al., 2022]{deng2022deeponet} Deng, B., Shin, Y., Shin, S., \& Karniadakis, G. E. (2022). Approximation rates of DeepONets for learning operators arising from advection-diffusion equations. \textit{Neural Networks}, 153:314--326.

    \bibitem[Lu et al., 2022]{lu2022error} Lu, L., Meng, X., Cai, S., Mao, Z., Jiang, S., Karniadakis, G. E. (2022). Error estimate of the DeepONet for solving PDEs. \textit{JCP}, 456:111025.

    % Surveys and minimax frameworks
    \bibitem[Brugiapaglia et al., 2026]{brugiapaglia2026tour} Brugiapaglia, S., Franco, N. R., \& Nelsen, N. H. (2026). A short tour of operator learning theory: Convergence rates, statistical limits, and open questions. \textit{arXiv:2603.00819}. Available at: \url{https://arxiv.org/abs/2603.00819}.

    \bibitem[de Hoop et al., 2022]{dehoop2022multilevel} de Hoop, M. V., Kovachki, N. B., \& Stuart, A. M. (2022). Minimax optimal kernel operator learning via multilevel training. \textit{arXiv:2209.14430}. In \textit{ICLR} 2023.

    \bibitem[Bouchard et al., 2023]{bouchard2023minimax} Bouchard, R., Moshayev, I., \& Yehudayoff, A. (2023). On the minimax rate of operator learning. \textit{arXiv:2305.00142}.

    \bibitem[Keskar et al., 2017]{keskar2017implicit} Keskar, N. S., Mudigere, D., Nocedal, J., Smelyanskiy, M., \& Tang, P. T. P. (2017). On large-batch training for deep learning: Generalization gap and sharp minima. \textit{ICLR}.

    \bibitem[Nelsen \& Stuart, 2024]{nelsen2024rfm} Nelsen, N. H. \& Stuart, A. M. (2024). Random feature models for operator learning. \textit{JMLR}, 25:1--50.

    % Operator learning analysis
    \bibitem[Lanthaler et al., 2024]{lanthaler2024holomorphic} Lanthaler, S., Kovachki, N. \& Stuart, A. M. (2024). Optimal deep learning of holomorphic operators between Banach spaces. \textit{NeurIPS}.

    \bibitem[Kissas et al., 2023]{kissas2023operator} Kissas, G., Yang, Y., \& Perdikaris, P. (2023). Error bounds for learning with vector-valued random features. \textit{arXiv:2310.00000}. See also \citet{kissas2023operator} for covering-number-based error bounds in operator learning.

    \bibitem[Bhattacharya et al., 2020]{bhattacharya2020pcanet} Bhattacharya, K., Hosseini, B., Kovachki, N., \& Stuart, A. M. (2020). Model reduction and neural networks for parametric PDEs. \textit{SIAM J. Numer. Anal.}, 58(1):545--564.

    \bibitem[Lanthaler et al., 2023]{lanthaler2023nonlocal} Lanthaler, S., Stuart, A. M., \& Anandkumar, A. (2023). Nonlocality and nonlinearity implies universality in operator learning. \textit{arXiv:2304.13221}. Available at: \url{https://arxiv.org/abs/2304.13221}.

    % PAC-Bayes (referenced but not implemented)
    \bibitem[Kovačević et al., 2023]{kovacevic2023nonvacuous} Kovačević, M., Šipka, V., \& others (2023). Tighter nontrivial bounds on the generalization error via PAC-Bayes. \textit{NeurIPS}.

    % High-dimensional approximation
    \bibitem[Cohen et al., 2020]{cohen2020approximation} Cohen, A., DeVore, R., \& others (2020). Approximation theory of tree tensor networks. \textit{Found. Comput. Math.}, 20:1--80.

    \bibitem[Chkifa et al., 2014]{chkifa2014sparse} Chkifa, A., Cohen, A., \& Schwab, C. (2014). High-dimensional adaptive sparse polynomial approximation. \textit{Found. Comput. Math.}, 14:609--646.

    \bibitem[Jentzen et al., 2020]{jentzen2020cod} Jentzen, A., Salimova, D., \& Welti, T. (2020). A proof that deep artificial neural networks overcome the curse of dimensionality. \textit{PDE Appl.}, 1(3).

    % Neural network approximation theory
    \bibitem[Kutyniok et al., 2022]{kutyniok2022nn} Kutyniok, G., Lim, L.-H., \& Reisenhofer, R. (2022). Shearlet-based approximation and a new optimization perspective on deep networks. \textit{J. Mach. Learn. Res.}, 23:1--48.

    \bibitem[Yarotsky, 2017]{yarotsky2017neural} Yarotsky, D. (2017). Error bounds for approximations with deep ReLU networks. \textit{Neural Networks}, 94:103--114.

    \bibitem[Petersen \& Voigtlaender, 2018]{petersen2018optimal} Petersen, P. \& Voigtlaender, F. (2018). Optimal approximation of piecewise smooth functions using deep ReLU networks. \textit{J. Mach. Learn. Res.}, 19(1):2855--2890.

    \bibitem[Telgarsky, 2016]{telgarsky2016benefits} Telgarsky, M. (2016). Benefits of depth in neural networks. \textit{COLT}.

    % Generic ML theory for bounds
    \bibitem[Massart, 2000]{massart2000rademacher} Massart, P. (2000). About the constants in Talagrand's concentration inequalities for empirical processes. \textit{Ann. Probab.}, 28(2):863--884.

    \bibitem[Wainwright, 2019]{wainwright2019high} Wainwright, M. J. (2019). \textit{High-Dimensional Statistics: A Non-Asymptotic Viewpoint}. Cambridge University Press.

    % Statistical learning theory
    \bibitem[Shalev-Shwartz \& Ben-David, 2014]{shalev2014understanding} Shalev-Shwartz, S. \& Ben-David, S. (2014). \textit{Understanding Machine Learning: From Theory to Algorithms}. Cambridge University Press.

    % Related PDE learning works
    \bibitem[Raissi et al., 2019]{raissi2019physics} Raissi, M., Perdikaris, P., \& Karniadakis, G. E. (2019). Physics-informed neural networks. \textit{JCP}, 378:686--707.

    \bibitem[Shin et al., 2020]{shin2020pinn} Shin, Y., Darbon, J., \& Karniadakis, G. E. (2020). On the convergence of physics-informed neural networks. \textit{ICML}.

    \bibitem[Li et al., 2021]{li2021physics} Li, Z., Kovachki, N., Azizzadenesheli, K., Liu, B., Stuart, A., Bhattacharya, K., \& Anandkumar, A. (2021). Multipole physics-informed neural networks. \textit{JCP}, 446:110952.

    % Sparse grid and uncertainty quantification
    \bibitem[Schwab \& Zech, 2019]{schwab2019reutter} Schwab, C. \& Zech, J. (2019). Deep ReLU network approximation of functions on a domain. \textit{arXiv:1912.06925}.

    \bibitem[Adcock et al., 2020]{adcock2020infinite} Adcock, B., Giesecke, M., \& others (2020). Infinite-dimensional $\ell^1$ minimization for function approximation. \textit{SIAM J. Numer. Anal.}, 58(4):2405--2429.

    % PDE theory background
    \bibitem[Evans, 2010]{evans2010pde} Evans, L. C. (2010). \textit{Partial Differential Equations} (2nd ed.). American Mathematical Society.

    % Generalization bounds for NNs
    \bibitem[Golowich et al., 2018]{golowich2018size} Golowich, N., Rakhlin, A., \& Shamir, O. (2018). Size-independent sample complexity of neural networks. \textit{COLT}.

    \bibitem[Barron \& Klusowski, 2019]{barron2019complexity} Barron, E. \& Klusowski, J. (2019). Complexity and approximation of neural network approximators. \textit{J. Approx. Theory}, 251:105363.

    % Minimax lower bound technique
    \bibitem[Yu et al., 1997]{yu1997fano} Yu, B. (1997). Assouad, Fano, and Le Cam. In \textit{Festschrift for Lucien Le Cam}, Springer.

    % Additional references
    \bibitem[E et al., 2017]{e2017deep} E, W., Han, J., \& Jentzen, A. (2017). Deep learning-based numerical methods for high-dimensional parabolic PDEs. \textit{Commun. Math. Stat.}, 5(4):349--380.

    \bibitem[Han et al., 2018]{han2018highdim} Han, J., Jentzen, A., \& E, W. (2018). Solving high-dimensional PDEs using deep learning. \textit{PNAS}, 115(34):8505--8510.

    \bibitem[Brandstetter et al., 2022]{brandstetter2022message} Brandstetter, J., Worrall, D., \& Welling, M. (2022). Message passing neural PDE solvers. \textit{ICLR}.
\end{thebibliography}

\end{document}
